### 2017-03-06 lafaySimulatedScenePerception.tex
rm density and diversity from analysis

% Instructions for Authors fuer Acta Acustica

\documentclass[twoside,twocolumn]{article}
\usepackage{acta}
\usepackage{url}
\usepackage[utf8]{inputenc}
\usepackage{setspace}

@@

% \footnotemark[1] \Authornotes{\footnotetext[1]{} }

@@

\Englishtitle{Investigating soundscapes perception through acoustic scenes simulation. Part I: simulation protocol presentation and case study} \Germanorfrenchtitle{}

@@

\Englishabstract{This paper introduces a new experimental protocol to study mental representations of urban soundscapes through a simulation process. Subjects are asked to create a full soundscape by means of a  software dedicated to sound \gl{edition}, coupled with a structured sound data set. This paradigm is used to characterize urban sound environment representations by analyzing the sound classes that were used to simulate the auditory scenes. A rating experiment of the soundscape pleasantness using a 7-points bipolar scale is conducted to further refine the analysis of the simulated urban acoustic scenes. Results show that 1) a semantic characterization in terms of presence~/ absence of sound sources is an effective way to characterize urban soundscapes pleasantness, and 2) physical descriptors computed for specific sound sources better characterize the appraisal than global descriptors.}
\Germanorfrenchabstract{}

@@

%\onecolumn
\setlength{\parindent}{5ex}
%\setstretch{2}
\section{Introduction}

%% Original intro

%% The notion of soundscape has been introduced by Schafer \cite{schafer_new_1969, schafer1977tuning} in the 1970s as the auditory equivalent to landscape. Following this paradigm, a sonic environment is described by focusing on the listener's evaluation, rather than only taking into account the acoustic parameters of the sound.

%% Only during the 1980s did policy-makers started taking into account the link between \textit{noise} and pollution, considering noise as a significant degradation of the quality of life. To fight this phenomenon, the first approach consisted in identifying unwanted sounds, and lowering their intensities. Thus, in the wake of this realization, several regulations have emerged, often limited to enforcing sound level thresholds.

%% However, it is now widely accepted that acoustic comfort is a complex concept that cannot be described solely with the help of objective acoustical measures such as $LA_{(eq,V)}$, but needs to take into account other perceptive factors \cite{schulte-fortkamp_soundscape_2006, aletta2016soundscape, Yang2005211, kang_semantic_2010}. ``\,Noise\,''  is a cognitive object which depends upon a listener's appreciation and the context in which it is heard. Many urban ``\,noises\,'', such as that of a siren, can annoy as well as warn of a danger, and many town districts are appreciated because of a lively and animated atmosphere which often results in higher sound levels. Conventional noise regulations concentrate on high sound levels from transportation and industry, and their scope does not extend beyond unwanted sound sources. In other words, they do not try to improve the sonic environment by identifying and reinforcing pleasant or important sounds. This latter approach was pioneered by Schafer, who named it the ``\,positive\,'' approach.

%% For two decades, the soundscape paradigm has been considered to study the perception of acoustic scenes in order to provide urban planners with meaningful sound quality indicators (see \cite{dubois2006cognitive} and  \cite{aletta2016soundscape} for extensive reviews). Most studies follow the \emph{modus operandi} of cognitive psychology \cite{maffiolo_caracterisation_1999}, by first describing the soundscape using both acoustic measurements and perceptive factors, and then investigating the interactions between those descriptors to understand the cognitive processes involved in the perception of the acoustic scenes. Soundscape studies are thus interdisciplinary \cite{davies_perception_2013, aletta2016soundscape}, borrowing tools from various fields such as acoustics, psycho-physics or cognitive psychology, but also psycho-linguistics, sociology and data mining.

%% The soundscape approach appears to be a powerful tool to develop perceptively motivated indicators \cite{schulte-fortkamp_soundscape:_2007}. That said, the use of various experimental protocols to assess and measure soundscapes makes the integration of results difficult \cite{davies_perception_2013}. Also, there still exist no definite consensus in the soundscape community about the acoustic descriptors to be used \cite{aletta2016soundscape}. This issue prevents experimenters to provide  urban planners with generic and clear indicators or models able to predict sound environment quality \cite{schulte-fortkamp_soundscape:_2007, aletta2016soundscape}. Recently, several projects such as the European Cooperation in Science and Technology Action (TD0804: soundscape of European Cities and Landscapes) or the Positive Soundscape project \cite{davies_perception_2013} have been undertaken to standardize soundscape assessment and descriptors \cite{schulte2013soundscape}, but those matters are still considered as open problems.

%% We believe that soundscapes studies greatly suffer from one important issue: to the best of our knowledge, there is no morphological definition of a soundscape, and therefore no standardized way to decompose it in distinct elements. It is thus difficult to study precisely which component of the environment is the cause of potential variations in the perceptive measurements.

%% We propose to tackle this issue by introducing a new experimental protocol, in which  subjects are asked to create a complex sonic environment thanks to a dedicated software tool linked to a dataset of isolated environmental sound samples. The simulation provides us with simulated scenes fully annotated in terms of source presence and sound levels. Those annotations enable us to ground the analyses of further perceptive descriptors obtained via standard perceptive evaluations of the simulated scenes, as well as to study the contribution of each sound source.

%% As a case study, we focus in this paper on urban soundscape pleasantness. We first give an overview of the tools used in soundscape perception studies, and justify the benefits of conducting this kind of simulation experiment as a data generation process. We then describe the simulation process, and present the results of the case study. The latter consists of two experiments: 1) a simulation experiment where subjects generate both ideal and non-ideal urban sonic environments, and 2) a pleasantness evaluation of the simulated scenes using a 7 points bipolar semantic scale.

%% Results show that 1) outcomes identified in the literature \cite{guastavino_categorization_2007} can be achieved by this approach, providing evidence that it has solid grounding,  2) sound markers that greatly affect the pleasantness of the soundscape can be identified with a good level of categorical detail, and 3) that physical descriptors computed specifically on the sound markers better characterise the appraisal than global descriptors.

%% Version MR 19/05/2016

The study of our sonic environments, both natural and man-made, has gathered growing interest in the past decades, finding applications in fields as varied as context inference and surveillance \cite{heittola13,1540194,park14}, evaluation of urban noise pollution \cite{raimbault_urban_2005} or the emerging area of eco-acoustics \cite{ECOACOUSTICS2014, krause}. The notion of \emph{soundscape}, introduced by Schafer \cite{schafer_new_1969, schafer1977tuning} in the 1970s as the auditory equivalent to landscape, is a theoretical framework of choice for that research. Following this paradigm, a sonic environment is described by focusing on the listener's evaluation \gl{and perception}, rather than only taking into account the acoustic parameters of the sound scene.

Most soundscape studies follow the \emph{modus operandi} of cognitive psychology \cite{maffiolo_caracterisation_1999}, by first describing the soundscape using both acoustic measurements and perceptual factors, and then investigating the interactions between those descriptors to understand the cognitive processes involved in the perception of the acoustic scenes. Soundscape studies are thus interdisciplinary \cite{davies_perception_2013, aletta2016soundscape}, borrowing tools from various fields such as acoustics, psychophysics or cognitive psychology, but also psycho-linguistics, sociology and data mining.

While the soundscape approach appears to be a powerful tool to develop perceptively motivated indicators \cite{schulte-fortkamp_soundscape:_2007}, the use of various experimental protocols to assess and measure soundscapes makes the integration of results difficult \cite{davies_perception_2013}. Moreover, there still exist no definite consensus in the soundscape community about the acoustic descriptors to be used \cite{brocolini2012prediction,aletta2016soundscape}. We believe that soundscapes studies greatly suffer from one important issue: to the best of our knowledge, there is no morphological definition of a soundscape, and therefore no standardized way to decompose it into distinct elements. It is thus difficult to study precisely which component of the environment is the cause of potential variations in the perceptual measurements.

We propose to tackle this issue by introducing a new experimental protocol, in which subjects are asked to create a complex sonic environment thanks to a dedicated software tool linked to a dataset of isolated environmental sound samples. The simulation provides us with simulated scenes fully annotated in terms of source presence and sound levels. Those annotations enable us to ground the analyses of further perceptual descriptors obtained via standard perceptual evaluations of the simulated scenes, as well as to study the contribution of each sound source.

As a case study to demonstrate the interest of that new approach, we focus in this paper on the issue of urban soundscape pleasantness as tackled in \cite{guastavino_ideal_2006} \nm{dans [14], il est plutot question de ideal / non ideal urban soundscapes ..... }.

Only during the 1980s did policy-makers start taking into account the link between \gl{``\,community noise\,''} and pollution, considering noise as a significant degradation of the life quality. To fight this phenomenon, the first approach consisted in identifying unwanted sounds, and lowering their intensities. Thus, in the wake of this realization, several regulations have emerged, often limited to enforcing sound level thresholds.

However, it is now widely accepted that acoustic comfort is a complex concept that cannot be described solely with the help of objective acoustical measures such as $L_{Aeq}$, but needs to take into account other perceptual factors \cite{schulte-fortkamp_soundscape_2006, aletta2016soundscape, Yang2005211, kang_semantic_2010}. ``\,Noise\,''  is a cognitive object which depends upon a listener's \gl{assessment} and the context in which it is heard. Many urban ``\,noises\,'', such as that of a siren, can annoy as well as warn of a danger, and many town districts are appreciated because of a lively and animated atmosphere which often results in higher sound levels. Conventional noise regulations concentrate on high sound levels from transportation and industry, and their scope does not extend beyond unwanted sound sources. In other words, they do not try to improve the sonic environment by identifying and reinforcing pleasant or important sounds. This latter approach was pioneered by Schafer, who named it the ``\,positive\,'' approach.

The soundscape paradigm has been employed for two decades to study the perception of urban acoustic scenes (see \cite{kang2006urban,dubois2006cognitive,aletta2016soundscape} for extensive reviews), however, due to the limitations we have pointed out, experiments have failed to provide  urban planners with generic and clear indicators or models able to predict sound environment quality \cite{schulte-fortkamp_soundscape:_2007, aletta2016soundscape}. Recently, several projects such as the European Cooperation in Science and Technology Action (TD0804: soundscape of European Cities and Landscapes) or the Positive Soundscape project \cite{davies_perception_2013} have been undertaken to standardize soundscape assessment and descriptors \cite{schulte2013soundscape}, but those matters are still considered as open problems \cite{ribeiro2013heart}. We explore in this paper what additional insights our proposed ``\,generative\,'' approach to soundscape perception analysis can bring to light.

We first give an overview of the tools used in soundscape perception studies, and justify the benefits of conducting this kind of simulation experiment as a data generation process. We then describe the simulation process, and present the results of the case study. The latter consists of two experiments: 1) a simulation experiment where subjects generate both ideal and non-ideal urban sonic environments, and 2) a pleasantness evaluation of the simulated scenes using a 7 points bipolar semantic scale. \nm{en revoyant cette description, je me dis qu'il y a pt-etre un petit probleme methodologique dans la mesure ou l'on contruit les scenes selon un attribut semantique (l'idealite) et on les evalue sur une autre echelle (l'agrement) .....
Je pense qu'il faudra bien argumenter c point plus bas dans le texte}

Results show that \nm{Cette section sur les resultats est etonnemment placee, et a priori, de maniere pas tres orthodoxe. Je verrai plutot cette synthese des resultats dans l'abstract ou dans la conclusion ....} 1) outcomes identified in the literature \cite{guastavino_categorization_2007} can be achieved by this approach, providing evidence that it has solid grounding, 2) sound markers that greatly affect the pleasantness of the soundscape can be identified with a good level of categorical detail, and 3) that physical descriptors computed specifically on the sound markers better characterize the appraisal than global descriptors.

\section{Background and motivations}
\label{statoftheart}

Two approaches, named in this paper \textit{categorical} and \textit{dimensional}, are commonly used to gain knowledge on soundscape perception \nm{this is not really specific to soundscape ... but a global approach of perception mechanisms. We can find this distinction also, for instance, in the works about musical timbre}.

The \textit{categorical} approach investigates both the existence of 1) soundscape categories and 2) soundscape components (categories of sound sources) using both sorting \cite{maffiolo_caracterisation_1999, guastavino_categorization_2007} and descriptive \cite{axelsson2005soundscape, raimbault_urban_2005, raimbault2006qualitative, guastavino_ideal_2006} tasks (questionnaires or interviews). To identify categories, those methods usually require psycho-linguistic and lexical analyses. Their results tend to show that soundscape perception relies upon the identification and assessment of the sound sources which compose the soundscape \cite{dubois2006cognitive}. Maffiolo \cite{maffiolo_caracterisation_1999} shows the existence of two generic soundscape categories, namely the ``\,event sequences\,'' and the ``\,amorphous sequences\,''. \gl{This distinction is also observed in \cite{guastavino_ideal_2006}}. Guastavino \cite{guastavino_categorization_2007} as well as Niessen \emph{et al.} \cite{niessen_categories_2010}  show the prominence of specific sound source categories in soundscape perception: ``\,natural\,'', ``\,human\,'', ``\,traffic\,'' and ``\,mechanical\,''. Other studies not using the \textit{categorical} approach reveal that sound source identification underlines the perception of others factors, for example Kuwano \emph{et al.} \cite{kuwano_memory_2003-1} show that overall judgment of loudness of a sound environment sequence is not statistically different from average instantaneous judgment of the recalled sound events of the sequence \nm{Cet exemple n'est pas completement clair pour moi :
certes, la loudness est bien une dimension continue (et donc en cela rentre bien dans ta terminologie de depart) mais je ne vois pas trop en quoi elle a un rapport avec la question de l'identification des sources ...
a preciser, je pense.}.  As reported by \cite{davies_perception_2013}, while the categorical approach does help to understand how a soundscape is composed, it does not provide information on how soundscape components are perceived and influence the qualitative evaluation.

Instead of asking subjects to break soundscapes down into categories, the \textit{dimensional} approach makes use of bipolar semantic scales to describe the soundscape either in a holistic way \cite{jeon2013soundwalk, kang_semantic_2010, cain_development_2013, torija2013application} or by considering the influence of specific sound sources \cite{guyot2005urban, lavandier2006contribution, ricciardi2015sound}. From those descriptors, it is possible to use dimensional analysis techniques such as Principal Component Analysis to find relevant perceptual axes \cite{cain_development_2013}, or to connect those perceptive outcomes with objective acoustics measurements using various statistical analysis or modelling techniques such as multiple linear regression \cite{lavandier2006contribution,ricciardi2015sound} or artificial neural network \cite{yu2009modeling}. Several semantic scales show promising results when it comes to modeling perceived affective quality, namely ``\,Calmness\,'' and ``\,Vibrancy\,'' \cite{cain_development_2013, axelsson2010principal}, or ``\,Pleasantness\,'' and ``\,Eventfulness\,'' \cite{axelsson2010principal}; but it has been shown that others context dependent (``\,Appropriateness\,'',``\,Familiarity\,'') or music related (``\,Music-likeness\,'' ) scales provide useful complementary information on soundscape perception \cite{aletta2016soundscape}. In addition, several studies have shown that multiple factors such as physical, behavioral, social, demographical, and psychological are involved when evaluating acoustic comfort \cite{marquis2005noise, yu2009modeling}. \gl{In particular} \cite{yu2010factors} shows that age and education level significantly influence the sound preference \nm{Je pense qu'il faut etre un peu plus critique sur ces questions la (et celles developpees dans la phrase precedente) et sur ce sujet qu'il faut absolument citer le papier de synthese de C. Marquis-Favre :} \gl{Attention \cite{marquis2005noise} est une review ...}.

\gl{Recent} studies investigating the influence of sound sources in soundscape perception show that global sound level (perceived or measured) cannot entirely explain the quality of a sonic environment. \gl{voir remarque precedente; sur cette question, je privilegierais plutot un artcile de revue (AAA), voire un bouquin (Kang,2006) qu'un proc. de conf. ...} \cite{guyot2005urban} shows that a rise in soundscape quality between two street soundscapes having a similar time-ratio of traffic sounds and a similar global sound level may result from the addition of a positively connotated sound source, e.g. belonging to the ``\,human\,'' category \nm{mais pas que de la categorie human; cf. le resultat que tu evoques  plus bas de [13] avec les sources type "public transport" egalement connotes plus positif que des "vehicules legers" du fait de leur caractere ecologique et collectif plus important que la voiture individuelle} \gl{eg means exempli gratia (par exemple), de plus nous montrons que bien  que les bus soient acceptés (presence dans les i-scenes), ils ne sont pas caracteristiques d'un environnement ideal (presence dans les ni-scenes)}. Using multiple linear regression to predict unpleasantness from descriptors characterizing global sound level as well as time presence and sound level of each source, \cite{lavandier2006contribution} shows that the percentage of explained variance increases when the source descriptors are added to the global sound level. Similar findings are made in \cite{ricciardi2015sound}, showing that a combined use of global and source specific descriptors allows to build more relevant model of sound quality. Lastly, \cite{lavandier2006contribution} shows that global sound level is not always appropriate to characterize the appraisal, especially in the case of positively rated soundscapes such as \textit{park} or \textit{market}.

In order to be able to study the relative perceptual importance of each sound source, it is beneficial to first gain knowledge about the presence and level of those sound sources within \nm{inside ?} the scene. Acquiring such data may be a serious challenge. From most of the studies cited above \cite{guyot2005urban, lavandier2006contribution, ricciardi2015sound}, subjects are asked to evaluate the time presence and/or the sound level of specific sound source categories using bipolar scales. To the best of our knowledge, only \cite{lavandier2006contribution} relies on objective measurement of those indicators by manually annotating the recordings used as stimuli. This annotation step is hardly manageable for \emph{in-situ} studies such as \cite{ricciardi2015sound} involving a large number of participants. Furthermore, only a pre-imposed restricted set of sound categories are evaluated, preventing experimenters to investigate the influence of more precise or alternative organisations of the sound sources.

There is thus a need to 1) ground the study into a precise taxonomy of sound sources, and 2) generate stimuli that are precisely annotated in terms of sound source presence and level. We propose to tackle this issue through the use of a simulation protocol in which subjects are asked to simulate specific complex sound environments thanks to a soundscape simulator linked to a dataset of isolated environmental sound samples. This approach leads us to consider the soundscape as a sequence of sound sources \cite{nelken_ear_2013}, following a morphological approach of great potential coined as a ``\,skeleton of events on a bed of texture\,''. Experimental subjects are able to create a sound environment by controlling separately each of the sources that make it up. Consequently, detailed post-analysis measurement of the impact of each source on the global perception of the simulated sound environment can be made.

Previous works conducted by Bruce \textit{et al.} also use a soundscape simulator to study soundscape perception \cite{bruce2009development,bruce2014effects}. However, in their approach, subjects may only add or remove a reduced set of long duration recordings of foreground sounds, and adjust their levels. In our approach, subjects are presented with a full dataset of classes of isolated sound samples, that they may finely manipulate in terms of sound intensity, repetition and time positioning. Important questions about the representativity of the  dataset are studied in Sections~\ref{sec:SoundDataset} and  \ref{sec:datasetAnalyses}.

@@

\section{Simulation protocol}

\subsection{Sound dataset}
\label{sec:SoundDataset}

The proposed protocol can be seen as the inverse of that of a description task (see Figure~\ref{psycholing}). In contrast with interviews and questionnaires, which both require subjects to describe a soundscape, \textit{i.e.} to decompose it into elements, subjects are here asked to recompose the soundscape from a sound dataset of urban environmental sounds.

\gl{To simulate a verisimilar sound environment, we need to provide the subject with a reasonable and representative number of isolated sounds. This dataset represents the ``\,sonic world\,'', or more precisely a semantic discretization of it in terms of  ``\,sound sources categories\,''. The term ``\,category\,'' refers to a group of objects (in our case sound sources) considered as equivalent.}

\gl{The design of the dataset is largely inspired by early work by Rosch \cite{roach1978cognition} on the formation of categories of concrete objects. According to Rosch, the category system may be viewed as a two dimensional space, where the ``\,vertical\,'' dimension deals with the inclusiveness of the categories (\emph{e.g.} how are ordered ``\,human\,'', ``\,voice\,'' and ``\,laugh\,'') while the ``\,horizontal\,'' dimension concerns the segmentation of categories lying in the same level of inclusiveness (\emph{e.g.} ``\,voice\,'' and ``\,car\,''). The ``\,vertical\,'' dimension is a taxonomy, where categories are related to one another by means of class inclusion. Rosch formalized three levels of abstraction to structure the ``\,vertical\,'' dimension: superordinate, basic and subordinate. The greater the inclusiveness of a category the higher the abstraction level. Several studies focusing on sound classification  have already used such hierarchical system \cite{houix_lexical_2012}. Following these considerations, our dataset is also structured along two dimensions:}

\begin{itemize}
\item object dimension: where we make the difference between distinct sounds sources such as ``\,voice\,'', or ``\,car\,'' and which is closely related to the ``\,horizontal\,'' dimension of Rosh.
\item abstraction dimension: where we make the difference between distinct level of abstraction use to describe a sound (``\,human\,''>``\,voice\,''>``\,laugh\,''), and which is closely related to the ``\,vertical\,'' dimension of Rosh.
\end{itemize}

Figure~\ref{datasetexample} shows a subset of the described hierarchical organization. The top classes of our taxonomy represent concepts such as ``\,urban transport\,'', encompassing classes of sound sources such as ``\,boat\,'' or  ``\,cars\,'', which are themselves split into sub-classes. At the leaf level, classes are described using sound source names and actions (\textit{car passing, male-voice yelling}). The deeper the class level, the lower the variability between the exemplars lying under the class. In this paper the class levels will be referred to as \emph{semantic levels}, the top classes being at semantic level 0, the first subclasses at semantic level 1 and so on. Leaf classes are collections of recorded isolated sounds.

The dataset taxonomy being the same for each subject, it acts as a reference for the analysis, allowing us to get a non ambiguous identification of the sounds populating a scene. Most soundscape studies use recordings or \textit{in situ} sound environment as stimuli. In both cases, these stimuli remain specific exemplars of many potential soundscapes which could occur at the same location and the same time. The re-composition process frees experimenters from this bias as the simulated soundscape is directly related to the subject's mental representation. While expressivity in this approach is still limited by the diversity of the available sound dataset, we nonetheless believe that it allows the subjects' responses to be less constrained than by an imposed soundscape exemplar, provided that the sound dataset is carefully designed.

\gl{As stated is section \ref{statoftheart}}, a soundscape  is here understood as being  ``\,a skeleton of events on a bed of texture\,'' \cite{nelken_ear_2013}, where sound events are short isolated sounds such as \textit{car passing} or \textit{baby crying}, whereas sound textures are long stationary  sounds such as \textit{traffic} or \textit{crowd sound}, in which the physical characteristics remain stable over long periods of time~\cite{saint-arnaud_classification_1995}. Several studies point out the fact that textures and events drive two distinct cognitive processes \cite{maffiolo_caracterisation_1999, mcdermott2013summary}. We thus clearly distinguish between ``\,sound events\,'' and ``\,sound textures\,'' and use two distinct datasets.

For ease of reading, we introduce some terms that will be used throughout the paper:

\begin{description}
\item[Sample]: the recording of an isolated sound event or texture.
\item[Sound class]: either a collection of samples or a collection of sub-classes, depending of the semantic level of the class.
\item[Track]: a temporal sequence of samples belonging to the same sound class. As samples of a texture track are concatenated seamlessly (see section \ref{sec:simulationProcess}), a texture track can be seen as being made of one sample.
\end{description}

\begin{figure}[t]
\begin{center}
\includegraphics[width=.4\paperwidth]{../gfx/1.eps}
  \caption{\label{psycholing} Psycho-linguistic analysis paradigm (top) and proposed approach (bottom).}
  \end{center}
\end{figure}

\begin{figure}[t!]
\begin{center}
\includegraphics[width=.4\paperwidth]{../gfx/3.eps}
\caption{\label{datasetexample} Hierarchical structure of the event and texture data sets. Depending on the considered top class, there could be more or less than 3 semantic levels.}
\end{center}
\end{figure}

\subsection{Sample Selection process}

The sample selection process is designed to rely only on the listening of sounds themselves, in order not to influence subjects with \textit{a priori} associated semantic values. To explore the dataset without any written textual help, we have designed a selection interface where each leaf class is depicted by a point in 2D space. \gl{The interface performances are extensively tested in the following study} \cite{lafayJAES}. Subjects may only interact with the leaf classes of the datasets but the location of the points in space is constrained by the taxonomy, \textit{i.e.} subclasses belonging to a same class are close to one another. This organization is chosen to facilitate the sound class selection process and allow the subject to quickly apprehend the full range of the dataset. This may also reduce several potential biases of studies relying on verbal data to describe a soundscape: 1) the variable fluency of the subject in the experiment language, 2) the lack of definite terms to fully describe sounds or complex sound environments \cite{guastavino_ideal_2006}, and 3) the high inter-subject variability in the verbal description of a same sound.

\subsection{Simulation process and parameters}
\label{sec:simulationProcess}

Following the above described terminology, a soundscape is a sum of tracks. Each track is modeled by a temporal sequence of sound samples belonging to the same sound class. Thus, each track is related to one specific leaf sound class (see Section \ref{sec:SoundDataset}). To generate a soundscape, subjects may interact with the track parameters, but not with specific samples; for example, if a subject wishes to put a sequence of car sounds, he may specify how often (on average) cars pass by, and how loud they should be on average, but he cannot control each event individually. This allows to streamline the creation process, and keep the user focused on a global perception of the soundscape.

The simulation process involves two steps:

>>>>

\documentclass[twoside,twocolumn]{article}
\usepackage{acta}
\usepackage{url}
\usepackage[utf8]{inputenc}
\usepackage{setspace}
\usepackage{graphicx} 
\usepackage{subfig}
\usepackage{caption}
\usepackage{sectsty}

\allsectionsfont{\bfseries}

\newtheorem{mydef}{Définition}

@@

@@

\newcommand{\ie}{\emph{i.\,e.}}
\newcommand{\Ie}{\emph{I.\,e.}}
\newcommand{\cf}{cf.}
\newcommand{\Cf}{Cf.}
\newcommand{\vs}{\emph{vs.}}
\newcommand{\Vs}{\emph{Vs.}}
\newcommand{\al}{\emph{et~al.}}
\newcommand{\eg}{\emph{e.\,g.}}
\newcommand{\Eg}{\emph{E.\,g.}} 
\newcommand{\etc}{\emph{etc.}}
\newcommand{\myfloatalign}{\centering}

\Englishtitle{Investigating soundscapes perception through acoustic scenes simulation.} \Germanorfrenchtitle{}

@@

\Englishabstract{This paper introduces a new experimental protocol to study mental representations of urban soundscapes through a simulation process. Subjects are asked to create a full soundscape by means of a  software dedicated to sound edition, coupled with a structured sound data set. This paradigm is used to characterize urban sound environment representations by analyzing the sound classes that were used to simulate the auditory scenes. A rating experiment of the soundscape pleasantness using a 7-points bipolar scale is conducted to further refine the analysis of the simulated urban acoustic scenes. Results show that 1) a semantic characterization in terms of presence~/ absence of sound sources is an effective way to characterize urban soundscapes pleasantness, and 2) physical descriptors computed for specific sound sources better characterize the appraisal than global descriptors.}
\Germanorfrenchabstract{}

@@

%\onecolumn
\setlength{\parindent}{5ex}
%\setstretch{2}

\section{Introduction}
\label{sec:intro}

\gl{[TODO ...]} \\

Comme nous l'avons vu, la recherche sur les paysages sonores a besoin d'outils permettant d'analyser séparément les influences des différentes sources sur les qualités affectives de l'environnement. La simulation offre des possibilités intéressantes (\cf~Section~\ref{sec:ch4_modAnaSo}), car elle nous permet d'obtenir des scènes sonores dont nous connaissons tous les paramètres structuraux, en particulier les caractéristiques distinctes des différentes sources.

Pour démontrer l'intérêt de la simulation en analyse sensorielle, nous choisissons, comme cadre applicatif, le problème de l'agrément perçu dans les environnements sonores urbains.

Ce Papier présente les résultats d'une série d'expériences s'appuyant sur la simulation, et visant, chacune, à comprendre comment les différentes sources sonores qui composent une scène influent sur la perception de l'agrément:

\begin{enumerate}
\item \emph{expérience 1.a, simulation}: au cours de cette expérience, les sujets simulent les environnements qui serviront de stimuli pour les étapes suivantes;
\item \emph{expérience 1.b, évaluation de l'agrément}: les sujets doivent évaluer l'agrément des scènes simulées à partir d'une échelle sémantique;
\item \emph{expérience 2, évaluation de l'agrément après modification des scènes}: comme pour l'expérience précédente, les sujets doivent évaluer l'agrément des scènes simulées à partir d'une échelle sémantique. Cependant les scènes ont été modifiées, \ie~privées de certaines classes de sons identifiées comme ayant un impact sur l'agrément perçu; 
\end{enumerate}

Le papier comprend 5 sections. La première constitue notre introduction. La deuxième introduit l'outil de simulation \emph{SimScene}. La troisième reprend l'expérience de simulation à proprement parler, ainsi que l'épreuve \emph{évaluation de l'agrément}. La quatrième décrit l'épreuve \emph{évaluation de l'agrément après modification des scènes}. Nous concluons dans la dernière partie.

Précisons que l'expérience de simulation a fait l'objet d'une étude pilote dont les résultats sont présentés dans  \cite{lafay2013atiam,lafay2014new}. L'outil de simulation \emph{Simscene} est présenté quant à lui en détails dans \cite{rossignol2015simscene,lafay2016JAES}. Nous résumons ici ses fonctionnalités d'importance pour notre étude.

@@

\section{L'outil de simulation}
\label{sec:simscene}

\emph{Simscene} est un environnement de travail audio-numérique dont la première version a été développée dans le cadre du projet HOULE\footnote{Projet HOULE : \url{http://houle.ircam.fr/}}. Il permet de simuler des paysages sonores à partir d'un corpus de sons. Il est prévu pour fonctionner sur les navigateurs internet \emph{Chrome} et \emph{Firefox}. L'outil a été développé en javascript à l'aide de la bibliothèque \emph{angular.js}\footnote{\emph{angular.js} : \url{https://angularjs.org/}} et du standard \emph{web-audio}\footnote{\emph{web-audio} : \url{http://www.w3.org/TR/webaudio/}}. L'interface de sélection (\cf~Section~\ref{sec:simscene_ssf}) a été développée à l'aide de la bibliothèque \emph{D3.js} \cite{d32011}.

Le fonctionnement de \emph{Simscene} se rapproche de celui d'un séquenceur audio. Chaque utilisateur choisit une classe de sons via l'interface de sélection (\cf~Section~\ref{sec:simscene_ssf}). Une fois la classe de sons sélectionnée, une piste audio, liée à cette classe, est créée. L'utilisateur peut alors modifier certaines propriétés de la piste via un groupe de paramètres de contrôle propre à chacune (\cf~Section~\ref{sec:simscene_parametre}). Des champs de texte sont prévus afin de permettre à l'utilisateur 1) de nommer chaque piste, 2) de donner un titre à la scène simulée 3) de commenter la scène simulée.

\subsection{Banque de données de samples}

\subsubsection*{Typologie}
\label{sec:simscene_Typologie}

Afin de créer un corpus de classe de sons de référence pour la simulation, nous réalisons une typologie des sons environnementaux urbains.

Pour ce faire, une étude bibliographique est effectuée, afin d'identifier les sources et ambiances sonores les plus souvent citées dans les publications. Cette étude porte sur 16 articles ou thèses. Chacun d'eux traite de la manière dont nous discriminons les paysages sonores urbains. Il ressort que plusieurs approches sont possibles:

\begin{itemize}
\item 9 articles abordent le problème par une approche perceptive, soit en identifiant ou répertoriant des catégories de sources sonores, soit en étudiant l'impact de classes de sons spécifiques sur la perception de l'environnement: \cite{maffiolo_caracterisation_1999,raimbault2002simulation,guastavino_etude_2003,defreville2004aactivity,raimbault2005urban,dubois2006cognitive,devergie_relations_2006,guastavino2006ideal,niessen2010categories}
\item 3 articles proposent une classification morpho-typologique, divisant l’environnement sonore urbain en «~zones sonores~» possédant une identité acoustique forte, selon la configuration et la pratique du site: \cite{maffiolo_caracterisation_1999,beaumont2004pertinence,polack2008perceptive}
\item 2 articles répertorient et classifient les sources sonores d’un point de vue expert: \cite{leobon_analyse_1986,brown2011towards}
\end{itemize}

Les réactions à la musique étant trop subjectives, et les jugements esthétiques ne pouvant qu'altérer les données d'évaluation, nous choisissons, dans cette étude, de ne pas considérer les sources musicales de type musiciens de rue, radios de voitures, d'appartements, \etc.

\subsubsection*{Événements et Textures}
\label{sec:simscene_eventTexture}

\subsubsection*{Taxonomie}
\label{sec:simscene_getSample}

Nous nommons sample, un enregistrement d'un son isolé, qu'il s'agisse d'un événement ou d'une texture. Chaque classe est alors comprise comme une collection de samples jugés perceptivement équivalents.

Les classes de sons sont organisées suivant une structure hiérarchique, afin de former une taxonomie. La structure taxonomique s'inspire grandement de l'axe vertical de l'organisation catégorielle proposée par E. Rosch \cite{rosch1978cognition}, \ie~plus le niveau d'abstraction de la classe est élevé, plus la description de la classe est précise, et plus les sources sonores incluses dans cette classe sont semblables. Si le niveau d'abstraction d'une classe est tel que cette dernière possède des sous-classes, alors sa collection de samples est la somme des collections respectives de chacune des sous-classes.

Nous établissons deux taxonomies, une pour les événements (\cf~Figures~\ref{fig:taxonomieEventa} et~\ref{fig:taxonomieEventb}), une autre pour les textures (\cf~Figures~\ref{fig:taxonomieTexturea} et~\ref{fig:taxonomieTextureb}).

Pour les événements, nous considérons quatre niveaux d'abstraction allant des classes les plus globalisantes (niveau d'abstraction 0), aux classes les plus spécifiques (niveau d'abstraction 3). Pour les textures, nous ne considérons que trois niveaux d'abstraction.

Notons que la taxonomie des événements est assez proche de celle de \cite{Salamon14}, introduite ultérieurement à notre étude.

\subsubsection*{Acquisition des samples}
\label{sec:simscene_getSample}

Sur la base des taxonomies précédemment établies, 483 sons isolés ont été collectés, dont 381 événements, et 102 textures. Parmi ces samples, 332 ont été enregistrés, et 151 proviennent de différentes banque de sons (\emph{SoundIdeas}\footnote{\emph{SoundIdeas} : \url{ http://www.sound-ideas.com/}} et \emph{Universal SoundBank}\footnote{\emph{Universal SoundBank} : \url{http://www.universal-soundbank.com/}}).

Tous les enregistrements ont été effectués à l’aide d'un micro canon \emph{AT8035}\footnote{Micro canon \emph{AT8035} : \url{http://eu.audio-technica.com/fr/products/product.asp?catID=1&subID=6&prodID=1845}} relié à un enregistreur \emph{ZOOM H4n}\footnote{Enregistreur \emph{ZOOM H4n} : \url{http://www.zoom.co.jp/english/products/h4n/}}. L’utilisation du micro canon nous permet d’isoler les événements sonores du brouhaha urbain. Pour les textures, il nous permet d’éviter les événements sonores proches du preneur de son. Nous pouvons ainsi pointer des "zones sonores", en nous tenant à une certaine distance de ces dernières, afin de capter uniquement le brouhaha émanant de la zone ciblée.

Tous les sons ont été normalisés au même niveau $RMS$ de $-12$ $dB$ (FS), \ie. relativement à un niveau pleine échelle (\emph{relative to Full Scale}). Dans notre cas, ce niveau pleine échelle est de 1 Volt.

\subsection{Paramètres}
\label{sec:simscene_parametre}

L'outil de simulation ne propose pas d’interagir avec un sample en particulier, mais toujours avec une piste, \ie~une séquence de samples. Plusieurs paramètres permettent au sujet de contrôler la structure de chaque piste.
